Analysis

### 2022-05-15 manuscript/manuscript.Rmd
More writeup

CTCF (CCCTC-binding factor) is a versatile transctiprion regulators that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu]. <!--It was originally defined as a transcriptional repressor of the MYC oncogene [@Lobanenkov:1990vm][@Filippova:1996tv].--> Originally defined as a repressor/insulator [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].

CTCF binds to different DNA sequences through combinatorial use of 11-zinc fingers. While many functional CTCF-binding sites do not match a consensus motif [@Ohlsson:2010uy], a multi-part CTCF-binding motif has been defined [@Schmidt:2012uk]. For the majority of CTCF–DNA- binding events, the N-terminal zinc fingers interact with a 14-bp long M1 motif. Additionally, the C-terminal fingers interact with a shorter M2 sequence, creating a 34-bp-long M1+M2 motif.

Topologically Associating Domains (TADs) are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa]. CTCF binding, especially in convergent orientation, has long been known to be critical for loop formation [@Dixon:2012aa][@Oti:2016aa].

The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj]. Unfortunately, the database has not been updated since 2013 and its data are limited to old genome assemblies (e.g., hg19, mm9).

@@

Chang et al. provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl]. They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif (mm10 genome assembly).

>>>>

The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiprion regulator that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux]. Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].

Most of CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux].

CTCF localization is largely invariant across different cell types [@Kim:2007ux]. The goal of our package is to provide such lists of genome-wide CTCF binding sites generally applicable across cell types.

CTCF binds to different DNA sequences through combinatorial use of 11-zinc finger domains to $\sim50bp$ target sites that have remarkable sequence variation with binding motif length varying between 20 and 40bp [@Ohlsson:2001wu]. A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1. It was subsequently refined to include an additional 9bp motif (M2) [@Schmidt:2012uk]. For the majority of CTCF–DNA- binding events, the N-terminal zinc fingers interact with a 14-bp long M1 motif. Additionally, the C-terminal fingers interact with a shorter M2 sequence, creating a 34-bp-long M1+M2 motif. The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al. [@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding.

Position Weight Matrices (PWMs) [@Stormo:2000uw] for CTCF are available from multiple databases. JASPAR 2022 [@Castro-Mondragon:2022uv] and HOCOMOCO [@Kulakovskiy:2018vw] databases list one 19bp-long CTCF PWM ([MA0139.1](https://jaspar.genereg.net/matrix/MA0139.1/) and [CTCF_HUMAN.H11MO.0.A](https://hocomoco11.autosome.org/motif/CTCF_HUMAN.H11MO.0.A), respectively). Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific [CTCF](http://cisbp.ccbr.utoronto.ca/TFreport.php?searchTF=T094831_2.00) binding motifs, likely reflecting the variability of CTCF motifs without considering the M1+M2 combinatorial structure. The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al. [@Schmidt:2012uk], the Ren_20 motif identified by Kim et al. [@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al. [@Xie:2007wm].

Unfortunately, the database has not been updated since 2013 and its data are limited to old genome assemblies (e.g., hg19, mm9).

CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa]. Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Rao:2014aa][@Vietri-Rudan:2015tu], and the enrichment of CTCF and cohesin is observed at TAD boundaries  [@Beagan:2020wz][@Stilianoudakis:2020aa]. To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops

We utilized the PWMscan web tool [@Ambrosini:2018wv] because it has a large collection of PWMs (JASPAR, HOCOMOCO, SwissRegulon, UniPROBE, CIS-BP, data from Jomla, Isakova publications) and genomes (49 genome assemblies); its results show over 99% agreement with FIMO, HOMER, and other motif scanning tools.

@@

Chang et al. provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl]. They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif (mm10 genome assembly). It should be noted that mouse embryonic stem cells (ESCs) have thousands of CTCF-binding events that cannot be conserved in the human genome, as they are found in rodent- specific B2 repeat elements [@Bourque:2008ws].
